---
title: "R_Learn2025"
format: html
editor: visual
---

R4DataScience2025

**Chapter 3: Data Transformation**

```{r}
#| label: load-packages
#| echo: False
#library Chunk
library(tidyverse)
library(nycflights13)

#notation, packagename::functionname()
```

**NOTE:** When loading tidyverse, dplyr has conflicts with base version functions, `filter()` and `lag()`. If you want to use the base version, code the whole function as above.

3.1.2

```{r}
#| label:nycflights13
#| echo: False 

#using nycflights13::flights, 336,776 from nyc in 2013
#option1
View(flights)
#option 2
#print(flights,width = Inf)
#option 3
#glimpse(flights)
```

`flights` is a tibble, a special data frame used by tidyverse to avoid common problems data frames have. Ex) prints only the first few rows and only the columns that fit on one screen. You can use `View(flights)` or `print(flights,width = Inf)` for all columns of use `glimpse()` . Options that are not View(), describe the data type for each column.

| Data Abbrev. | Type               |
|--------------|--------------------|
| `<int>`      | integer            |
| `<dbl>`      | double(Real number |
| `<chr>`      | character          |
| `<dttm>`     | date-time          |

: Types of Data in `flights`

3.1.3 dplyr basics

**Note:** dplyr functions can be called "verbs"

Three things dplyr verbs will have in common

1.  First argument is always a data frame.
2.  the subsequent arguments typically describe which columns to operate on using variable names (without quotes).
3.  The output is always a new data frame

Use the pipe, `|>` ,to chain multiple verbs together. The pipe passes what is to its left to the right. `x |> f(y)` (= `f(x,y)` ) is pronounced x 'then' f(y)

3.2 Rows
